% Equation Module: CMB Power Spectrum Modifications
% Purpose: Unified framework corrections to acoustic peaks and ISW effect
% Chapter: 20 (Cosmological Applications)
% Auto-numbered equation

\begin{equation}\label{eq:cosmo_cmb_modifications}
\boxed{
\begin{aligned}
C_\ell^{TT,unified} &= C_\ell^{TT,\Lambda CDM}\left(1 + \Delta_\ell^{Aether} + \Delta_\ell^{Genesis} + \Delta_\ell^{Pais}\right) \\[10pt]
\Delta_\ell^{Aether} &= \frac{\kappa^2\phi_0^2}{M_P^2}\frac{\ell(\ell+1)}{2\pi}\exp\left(-\frac{\ell^2}{2\ell_{damp}^2}\right) \\[6pt]
\Delta_\ell^{Genesis} &= -\frac{g_{AG}^2|\mathcal{F}|^2}{k_{Planck}^2 r_*^2}\exp\left(-\frac{\ell}{2000}\right) \quad \text{(small-scale cutoff)} \\[6pt]
\Delta_\ell^{Pais} &= \frac{v_{rec}^2}{c^2}\frac{r_S^{rec}}{r_*}\delta_{\ell,low} \quad \text{(ISW modification)} \\[10pt]
\ell_{peak}^{(n)} &= n\pi\frac{r_*}{r_s}, \quad r_s = \int_0^{t_{rec}}\frac{c_s\,dt}{a(t)}, \quad c_s^2 = \frac{1}{3(1 + R)} \\[6pt]
\Delta r_s^{unified} &= r_s\left(\frac{\kappa\phi_0}{M_P} - \frac{g_{AG}|\mathcal{F}|^2 r_s^2}{L_P^2}\right) \sim 0.001\, r_s
\end{aligned}
}
\end{equation}

\noindent\textbf{Physical Interpretation:}
\begin{enumerate}
  \item \textbf{Aether scalar-photon coupling}: Scalar field $\phi$ couples to photons via $\mathcal{L} \sim \kappa\phi F_{\mu\nu}F^{\mu\nu}$, modifying photon propagation. This shifts acoustic peak heights and positions. Damping scale $\ell_{damp} \sim 1500$ from Silk damping. Correction: $\Delta C_\ell / C_\ell \sim 10^{-3}$ at $\ell \sim 1000$.

  \item \textbf{Genesis small-scale cutoff}: Nodespace discreteness at Planck scale provides UV regularization. Power spectrum exponentially suppressed for $\ell > 2000$ corresponding to $k > k_{Planck}$. This is a \textit{smoking gun signature} testable with CMB-S4 (target sensitivity at $\ell \sim 5000$).

  \item \textbf{Pais frame-dragging ISW}: Gravitomagnetic fields from rotating structures modify Integrated Sachs-Wolfe (ISW) effect at low $\ell < 50$. Effect scales as $(v/c)^2$ during recombination. For $v_{rec}/c \sim 10^{-5}$ and $r_S^{rec}/r_* \sim 10^{-3}$: $\Delta_\ell^{Pais} \sim 10^{-10}$ (negligible).
\end{enumerate}

\noindent\textbf{Sound Horizon Shift:}
The sound horizon at recombination $r_s$ determines acoustic peak spacing. Unified framework modifies sound speed via:
\begin{equation}
c_s^2 = \frac{1}{3(1 + R)}\left(1 + \frac{\kappa\phi_0}{M_P}\right)
\end{equation}
Predicted shift: $\Delta r_s / r_s \sim 10^{-3}$ (0.1\% effect, detectable with Planck).

\noindent\textbf{Testable Predictions:}
\begin{itemize}
  \item High-$\ell$ suppression from Genesis: Measure $C_\ell$ for $\ell > 2000$ with CMB-S4 (2030+)
  \item Scalar coupling from Aether: Cross-correlate CMB with large-scale structure
  \item Modified ISW from Pais: Cross-correlate low-$\ell$ CMB with galaxy surveys
\end{itemize}
