\section{Phase 3: Dynamic Systems and Universal Emergence}

Phase~3 interprets the Superforce as an engine for dynamical self-organisation.
Nodespaces, origami folds, and fractal recursions interact to form a scale-free
network that spans multiple dimensional regimes. The results summarised here
retain their speculative status until the underlying calculations are audited.

\subsection{Nodespace Stabilisation and Coupling}
\begin{align}
  S_{\text{nodespace}}
    &= \int \sqrt{-g}\,\mathcal{F}(x,t,D,z)\,\dd^{n}x,
    \label{eq:genesis:nodespace-action}\\
  T(z_i, z_j)
    &= \exp\!\left[-\alpha\,\frac{\lvert z_i - z_j\rvert}{\lambda}\right],
    \label{eq:genesis:nodespace-tunnel}
\end{align}
where \eqref{eq:genesis:nodespace-action} models formation via fractal
harmonics and \eqref{eq:genesis:nodespace-tunnel} describes resonant tunnelling
between modular coordinates $z_i$ and $z_j$. These expressions need derivations
and dimensional verification before acceptance.

\subsection{Origami Dynamics}
\begin{equation}
  \mathcal{T}_{\text{origami}} = \int \mathcal{G}(x,\theta)\,\dd^{D}x,
  \qquad
  \frac{\partial \mathcal{A}_{\text{origami}}}{\partial t}
    = \kappa \sin\!\left(\frac{\theta}{2}\right),
  \label{eq:genesis:origami}
\end{equation}
with folding angle $\theta$ and elasticity constant $\kappa$. Origami folds are
treated as gateways between nodespaces; inclusion in the main text will require
justification of boundary conditions and physical observables.

\subsection{Fractal Recursions}
\begin{equation}
  E_n = E_0\,r^{n},\qquad
  \frac{\dd^{\alpha} t}{\dd t^{\alpha}} = k\,t^{n},
  \label{eq:genesis:fractal}
\end{equation}
encoding energy cascades and fractional-time evolution. Parameters
$(E_0,r,k,\alpha)$ remain to be constrained by either simulations or empirical
targets.

\subsection{Emergent Phenomena}
\begin{align}
  \mathcal{R}_{\text{emergence}}
    &= \sum_{n=0}^{\infty}\sin\!\left(\frac{\pi z}{n}\right),\\
  \mathcal{O}_{\text{nodespace}}
    &= \min\!\left(\int \mathcal{F}(x,t,D,z)\,\dd x\right),\\
  C(x)
    &= \int \mathcal{G}(x,t,D,z)\,\ee^{\imag \nu t}\,\dd x.
\end{align}
These expressions attempt to model symmetry reinforcement, optimisation, and a
proposed “consciousness resonance.” Each must be clearly tagged as conjecture
unless physically grounded.

\subsection{Scale-Free Nodespace Network}
\begin{equation}
  P(k) \propto k^{-\gamma},
\end{equation}
with $\gamma$ defining the fractal exponent governing connectivity distribution.
Nodespaces at high $k$ act as resonance hubs; cross-checks with graph-theoretic
results are pending.

\paragraph{Open Work.}
\begin{itemize}
  \item Provide rigorous derivations or simulations for \eqref{eq:genesis:nodespace-action}–\eqref{eq:genesis:origami}.
  \item Separate speculative phenomena (e.g. resonance-driven consciousness) into
        clearly marked outlook sections.
  \item Align notation with other frameworks to avoid symbol overload.
\end{itemize}
